A full approximation scheme multilevel method \\ for nonlinear variational inequalities

### 2023-07-14 paper/fascd.tex
I think this addresses PF comments in Introduction

\begin{abstract}
We present the \emph{full approximation scheme constraint decomposition} (FAScd) multilevel method for solving variational inequalities (VIs). FAScd is a common extension of both the full approximation scheme (FAS) multigrid technique for nonlinear partial differential equations, due to A.~Brandt, and the constraint decomposition (CD) method introduced by X.-C.~Tai for VIs arising in optimization.  We extend the CD idea by recognizing and exploiting the telescoping nature of certain function space subset decompositions arising from multilevel mesh hierarchies.  Then, by applying strong smoothers, with work proportional to the number of unknowns on a finite element mesh level, we demonstrate that FAScd V-cycles exhibit excellent mesh-independent convergence rates and that FAScd F-cycles (full multigrid) are optimal solvers.  The example problems include unilateral and bi-obstacle VI problems with symmetric linear, nonsymmetric linear, and nonlinear differential operators.  Our examples demonstrate very good convergence even on problems where other multilevel VI solvers struggle.
%\pef{I want to say it's better than monotone multigrid. Is this too subtle/too much?}
% I vote for not saying this in the abstract.  It is hard to say which aspect of FAScd is more important to beating monotone multigrid: stronger smoother or improved up-smooth CD.  Probably the former, but for now we don't have any illuminating analysis.
\end{abstract}

@@

The constraint decomposition (CD) methods of Tai \cite{Tai2003} are designed to solve coercive variational inequality (VI) problems arising as the Karush--Kuhn--Tucker optimality conditions for the minimization of a convex functional over a convex set.  These methods have been shown to converge for multilevel and domain decompositions in a finite element (FE) context; in the multilevel case, the CD method has been shown to have almost grid-independent error bounds, and near-optimal complexity, for elliptic, linear obstacle problems \cite[Subsection 5.4]{Tai2003}; see also \cite[Theorem 4.6 and Algorithm 4.7]{GraeserKornhuber2009}.  In this paper we extend the multilevel CD method by replacing the coarser-level corrections with a full approximation scheme (FAS; \cite{Brandt1977,Bruneetal2015}) approach, and we apply this new solver to nonlinear VI problems which do not arise from minimization.
%\pef{Note I've changed ``coarse-level corrections'' to ``coarser-level corrections'', to imply we're not just operating only on the coarsest grids, but also on intermediate grids.}
% Cool.

Specifically, our \emph{full approximation scheme constraint decomposition} (FAScd) method extends the algorithms in \cite{GraeserKornhuber2009} and \cite{Tai2003} in the following directions:
\renewcommand{\labelenumi}{\emph{(\roman{enumi})}}
\begin{enumerate}
\item The FAS coarse corrections are adapted for general VI problems. The needed computations require neither an objective function nor linearity of the residual operator.
\item Box constraints, namely upper and lower obstacles, are permitted; see Section \ref{sec:femultilevel}.
\item We observe an asymmetry special to multilevel CD methods, namely that up-smoothing is more effective than down-smoothing because a ``telescoping'' subset property can be exploited on the up-smoothing side.
\item We implement V-cycles (Algorithm \ref{alg:fascd}) and F-cycles (full multigrid; Algorithm \ref{alg:fascd-fcycle}).
\item By using a Newton-type active set method as a smoother, with a fixed number of Krylov iterations per step so that the work of the smoother is proportional to the number of unknowns on the given level, we demonstrate that F-cycle FAScd is an optimal solver.
\pef{Clarify/limit which classes of problems we're claiming this for.}
\end{enumerate}

\pef{I think we need some text here hitting the nail on the head about what's different to monotone multigrid methods (even if it's just the absence of the invasive modifications of the basis functions).}

In our experiments, FAScd V-cycles exhibit essentially the same mesh-independent convergence rates for VI problems that FAS multigrid V-cycles exhibit for the corresponding (unconstrained) PDE problems.  The F-cycle results in Section \ref{sec:results} show nearly-textbook multigrid efficiency, namely solutions within discretization error in work comparable to few smoother sweeps on the finest-level mesh \cite{BrandtLivne2011}, for classical Laplacian obstacle problems, $p$-Laplacian obstacle problems, and nonnegativity-constrained advection-diffusion equations.  Furthermore, these results are essentially insensitive to the length or structure of the free boundary, i.e.~the geometric complexity of the coincidence set.

The iterates from FAScd are admissible, and thus the nonlinear operator in the VI needs only be defined for admissible states.  For example, we solve an ice sheet geometry problem (Examples \ref{ex:sia}, \ref{ex:results:sia}) in which the ice flow formulas are only meaningful for surface elevation iterates that do not penetrate the bedrock\footnote{More generally, fluid-layer problems, with a non-negative thickness constraint, often require the solution of an auxiliary PDE on a domain determined inside the VI residual evaluation \cite{Bueler2021conservation}.}.  Note that non-admissible methods, including semi-smooth methods \cite{BensonMunson2006}, would require unnatural and invasive modifications of the operator formula in such a fluid-layer context.  FAScd solves this low-regularity ice sheet problem, wherein the solution has unbounded gradient at the free boundary, with a mesh-independent number of F-cycle iterations.

This paper is organized as follows.  In Section \ref{sec:vi} we recall the theory of coercive and monotone variational inequalities, with several example problems.  Section \ref{sec:cd} reviews the basics of CDs, as proposed by Tai \cite{Tai2003}, and their induced iterations.  Section \ref{sec:femultilevel} sets up multilevel FE hierarchies, and then Section \ref{sec:cdmultilevel} presents defect constraints generated via monotone injection operators \cite{GraeserKornhuber2009}.  Here we propose an apparently new understanding of telescoping sets for up-smoothing, based on an ``incomplete'' CD.  Section \ref{sec:vcycle} then states the FAScd V-cycle, with Section \ref{sec:implementation} addressing convergence criteria, F-cycles, and our smoother choice.  Section \ref{sec:results} gives FAScd numerical results from a parallel Firedrake \cite{Rathgeberetal2016} implementation for the problems introduced in Section \ref{sec:vi}.  The work concludes with discussion and outlook in Section \ref{sec:discussion}, and an appendix on how FAScd reduces to classical FAS multigrid for PDEs.

@@

FIXME the observation that up-smoothing is more efficient seems to be new; compare the comments on V(1,0) and V(1,1) cycles in \cite{GraeserKornhuber2009,Tai2003}

>>>>

\begin{abstract}
We present the \emph{full approximation scheme constraint decomposition} (FAScd) multilevel method for solving variational inequalities (VIs).  FAScd is a common extension of the full approximation scheme (FAS) multigrid technique for nonlinear partial differential equations, due to A.~Brandt, and the constraint decomposition (CD) method introduced by X.-C.~Tai for VIs arising in optimization.  We extend the CD idea by exploiting the telescoping nature of certain function space subset decompositions arising from multilevel mesh hierarchies.  Then, by applying strong smoothers, with work proportional to the number of unknowns on a finite element mesh level, we demonstrate that FAScd V-cycles exhibit excellent mesh-independent convergence rates and that FAScd F-cycles (full multigrid) are optimal solvers.  The example problems include unilateral and bi-obstacle VI problems, with differential operators which are symmetric linear, nonsymmetric linear, and nonlinear.  Our examples demonstrate very good convergence even on problems where other multilevel VI solvers struggle.
\end{abstract}

%\pef{I want to say it's better than monotone multigrid. Is this too subtle/too much?}
% I vote for not saying this in the abstract.  It is hard to say which aspect of FAScd is more important to beating monotone multigrid: stronger smoother or improved up-smooth CD.  Probably the former, but for now we don't have any illuminating analysis.

@@

The constraint decomposition (CD) methods of Tai \cite{Tai2003} are designed to solve coercive variational inequality (VI) problems arising as the Karush--Kuhn--Tucker optimality conditions for the minimization of a convex functional over a convex set.  These methods were shown to converge for multilevel and domain decompositions in a finite element (FE) context.  In the multilevel case, the CD method has almost grid-independent error bounds, and near-optimal complexity, for elliptic, linear obstacle problems \cite[Subsection 5.4]{Tai2003}, \cite[Theorem 4.6 and Algorithm 4.7]{GraeserKornhuber2009}.  In this paper we extend Tai's multilevel CD method by replacing the coarser-level corrections with a full approximation scheme (FAS; \cite{Brandt1977,Bruneetal2015}) approach, and we apply this new solver to a variety of general, i.e.~nonlinear and non-optimization, VI problems.

Specifically, our \emph{full approximation scheme constraint decomposition} (FAScd) method extends the algorithms in \cite{GraeserKornhuber2009} and \cite{Tai2003} in the following five directions:
\renewcommand{\labelenumi}{\emph{(\roman{enumi})}}
\begin{enumerate}
\item The FAS coarse corrections are adapted for general VI problems. The needed computations require neither an objective function nor linearity of the residual operator.
\item Box constraints, namely upper and lower obstacles, are permitted.
\item We implement V-cycles (Algorithm \ref{alg:fascd}) and F-cycles (full multigrid; Algorithm \ref{alg:fascd-fcycle}).
\item We identify a ``telescoping'' subset property which can be exploited on the up-smoothing side of a V-cycle for greater performance.
\item By using a Newton-type active set method as a smoother, with a fixed number of (Newton and Krylov) iterations per step, so that the work of the smoother is proportional to the number of unknowns on the given level, we demonstrate that F-cycle FAScd is an optimal solver for the examples tested (Section \ref{sec:results}).
\end{enumerate}

Prior multilevel algorithms for VIs have various substantial limitations relative to the FAScd framework.  The projected full approximation scheme (PFAS) algorithm of Brandt \& Cryer \cite{BrandtCryer1983} is apparently limited to linear complementarity problems and projected (truncated) Gauss-Seidel smoothers.  While FAScd reduces to general, nonlinear FAS for PDEs when the inequality constraints are removed (Appendix \ref{app:reductions}), PFAS apparently reduces to linear multigrid.  The multilevel CD iteration of Tai \cite{Tai2003} is formulated in terms of an objective function, so it only applies (as stated) to optimization-type problems, and furthermore each of its coarser-level problems refers to the finest-level discretization.  The problematic latter property is corrected by reformulation using defect constraints and monotone injection operators (see \cite[Algorithm 4.7]{GraeserKornhuber2009} and Section \ref{sec:cdmultilevel} herein).  The monotone multigrid methods of Kornhuber and co-authors \cite{GraeserKornhuber2009,Kornhuber1994} appear to be implemented only with a Gauss-Seidel type smoother, and applied only to optimization-type VI problems \cite[for example]{JouvetGraeser2013}.  Furthermore, monotone multigrid methods are more difficult to implement in FE libraries because modifications of the FE basis functions are needed to approximate the coarser-level coincidence sets.  No such invasive basis modifications are used either in Tai's original multilevel CD or in FAScd.  Our Firedrake library \cite{Rathgeberetal2016} implementation of FAScd requires no modifications of basis functions and yet it exhibits excellent performance, including on a problem used to demonstrate the benefits of monotone multigrid (Example \ref{ex:results:classical}; compare \cite[problem 7.1.1]{GraeserKornhuber2009}).

%\pef{I think we need some text here hitting the nail on the head about what's different to monotone multigrid methods (even if it's just the absence of the invasive modifications of the basis functions).}
% Gave it a shot.

In our experiments, FAScd V-cycles exhibit essentially the same mesh-independent convergence rates for VI problems that FAS multigrid V-cycles exhibit for the corresponding (unconstrained) PDE problems.  The F-cycle results in Section \ref{sec:results} show nearly-textbook multigrid efficiency, namely solutions within discretization error in work comparable to few smoother sweeps on the finest-level mesh \cite{BrandtLivne2011}.  Specifically we see this performance for classical Laplacian obstacle problems, $p$-Laplacian obstacle problems, and nonnegativity-constrained advection-diffusion equations.  Furthermore, these results appear to be insensitive to the length or structure of the free boundary, i.e.~the geometric complexity of the coincidence set.

All iterates in a FAScd cycle are admissible, and thus the nonlinear operator in the VI needs only be defined for admissible states.  For example, we solve an ice sheet geometry problem (Examples \ref{ex:sia} and \ref{ex:results:sia}) in which the ice flow formulas are only meaningful for surface elevation iterates that do not penetrate the bedrock.  (More generally, fluid layers subject to a non-negative thickness constraint typically require the solution of an auxiliary PDE on a domain determined inside the VI residual evaluation \cite{Bueler2021conservation}, but this is not needed in the ice sheet problem here.)  Non-admissible methods, including semi-smooth methods \cite{BensonMunson2006}, would require unnatural modifications of the operator formula.  FAScd solves this low-regularity ice sheet problem, wherein the solution has unbounded gradient at the free boundary, by a mesh-independent number of F-cycle iterations.

The paper is organized as follows.  In Section \ref{sec:vi} we recall the theory of coercive and monotone variational inequalities, with several example problems.  Section \ref{sec:cd} reviews the basics of CDs and their associated iterations.  Section \ref{sec:femultilevel} sets up multilevel FE hierarchies, and then Section \ref{sec:cdmultilevel} presents defect constraints generated via monotone injection operators \cite{GraeserKornhuber2009}.  Here we propose an apparently new understanding of telescoping sets for up-smoothing, based on an ``incomplete'' CD and an associated iteration.  Section \ref{sec:vcycle} then states the FAScd V-cycle, with Section \ref{sec:implementation} addressing convergence criteria, F-cycles, and smoother construction.  Section \ref{sec:results} gives numerical results, from a parallel Firedrake \cite{Rathgeberetal2016} implementation (see Code Availability), for each example problem introduced in Section \ref{sec:vi}.  The work concludes with a discussion and outlook in Section \ref{sec:discussion}.

@@

FIXME the observation that up-smoothing is more efficient, with consequent asymmetry between $V(1,0)$ and $V(0,1)$ performance, seems to be new; compare the comments on V(1,0) and V(1,1) cycles in \cite{GraeserKornhuber2009,Tai2003}
